External priors for the next generation of CMB experiments.

### 2015-09-17 Notes/N_eff_prior_paper.tex
minor edit notes

% Group addresses by affiliation; use superscriptaddress for long
% author lists, or if there are many overlapping affiliations.
% For Phys. Rev. appearance, change preprint to twocolumn.
% Choose pra, prb, prc, prd, pre, prl, prstab, prstper, or rmp for journal
%  Add 'draft' option to mark overfull boxes with black boxes
%  Add 'showpacs' option to make PACS codes appear
%  Add 'showkeys' option to make keywords appear
\documentclass[aps,prd,reprint,superscriptaddress]{revtex4-1}
%\documentclass[aps,prl,preprint,superscriptaddress]{revtex4-1}
%\documentclass[aps,prl,reprint,groupedaddress]{revtex4-1}
%\usepackage{bm}
\usepackage{amsmath}
\usepackage{lineno}
\usepackage{graphicx}
%<<<<<<< HEAD
%\input{papers_packages_command._include_tex}

%\newcommand\refsec[1]{\S\ref{sec:#1}}
%\newcommand\refeq[1]{Eq.~(\ref{eqn:#1})}
%\input{papers_packages_command._include_tex}
\newcommand\refsec[1]{\S\ref{sec:#1}}
\newcommand\refeq[1]{Eq.~(\ref{eqn:#1})}
\newcommand{\refssec}[1]{section~\ref{subsec:#1}}
\newcommand{\reffig}[1]{Fig.~\ref{fig:#1}}
%\newcommand{\apj}{ApJ}
\newcommand{\physrep}{Physics Reports}
\newcommand{\aapr}{A\&A Rev.}
\newcommand{\apjl}{ApJL}
\newcommand{\apjs}{ApJS}
\newcommand{\aap}{A\&A}
\newcommand{\mnras}{MNRAS}
%\newcommand{\prd}{Phys. Rev. D}
\newcommand{\physrev}{Phys. Rev.}
\newcommand{\physrevlett}{Phys. Rev. Lett.}
\newcommand{\jcap}{{\em JCAP }}
\newcommand{\aj}{AJ }

%>>>>>>> 683231e77bc036bad411dd2244ca99280568d723
\bibliographystyle{apsrev4-1}

@@

\date{\today}
\begin{abstract}
The next generation of cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, Dark Matter and Dark Energy. 
Indeed the low level of noise, together with improved angular resolution, will drastically increase the signal to noise of the CMB polarized signal as well as the reconstructed lensing potential of high redshift large scale structure. 
Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments. Here, we examine quantitatively with a Fisher matrix approach, the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 experiment.
We find that CMB S4 
\end{abstract}

@@

{\bf A lot of this is standard and need not be included, which will make the paper much shorter: a good thing!}

@@

\subsection{Fisher Matrix formalism \label{subsec:fm-formalism}}
To estimate errors on cosmological parameters we follow the Bayes theorem that relates the likelihood to measure a set of data given the parameters of the model $\mathcal{L}(d|\theta)$, to what we want: the posterior probability of those parameters given the data, $\mathcal{L}(\theta|d)$.
These two are related by the prior probability of the parameters $P(\theta)$ through:
\begin{equation}
\mathcal{L}(\theta|d)\propto \mathcal{L}(d|\theta)P(\theta),
\end{equation}
where, as usual, we have neglected the probability of the data $P(d)$.
These general framework can be simplified in our case. We will focus not on an entirely general posterior distribution, but we will assume that the likelihoods are gaussian.
Furthermore we will not derive the global shape of the likelihood but we will obtain parameters errors by studying small perturbations around its maximum. These are the two basic assumptions of the Fisher-matrix approach.

@@

\subsection{CMB S4 experiments specifications.\label{subsec:cosmo-noise}}
In this subsection we describe the assumptions we made in calculating the elements that goes in \refeq{Fij_def2} and, in particular, the power spectra $C_{\ell}$ and the noise power $N_{\ell}$.
Firstly we parametrize our cosmology using a flat $\nu \Lambda$CDM universe. We allow a set different Dark Energy models by introducing the equations of state parameter $w$ as a varying parameter.
We chose our fiducial parameters following Table 2 of \textit{Planck} best fit \cite{planck-collaboration:2014g}, i.e. $\Omega_c h^2 = 0.12029$, $\Omega_b h^2 = 0.022068$, $A_s = 2.215\times10^{-9}$ at $k_0 = 0.05\ {\rm Mpc}^{-1}$, $n_s = 0.9624$, $\tau = 0.0925$, and $H_0 = 67.11$ km/s/Mpc. Regarding the $\Lambda$CDM extension, we chose a neutrino energy densities $\Omega_{\nu} h^2$=0.0009, which corresponds to $M_{\nu}$ $\simeq$ 85\ meV, a standard $N_{\rm eff}=3.046$ and a dark energy equation of state, $w=-1$.
Given the model, we use CAMB \cite{Lewis:1999bs} to compute the power spectra $C_{\ell}$ at the fiducial values and at those needed to compute derivatives using \refeq{deriv}. Notice that while we vary one parameter in \refeq{deriv} we keep all the others fixed with the exception of $\Omega_{\Lambda}$ which is always changed in order to keep the universe flat ($\Omega_{\rm k} = 0$).

@@

\section{Results \label{sec:results}}
In this section we present our main results. We will start from

{\bf Let's open up one parameter at a time: first standard LCDM with neff, and then with mnu, and then with w. Also, let's show a curve in each plot with constraints on *all*  parameters, not just one of them, as that will give people a sense of what is gained if multiple priors can be obtained.}

@@

In the standard cosmology, three active neutrinos are thermally produced in the early universe. Were they to decouple well before the epoch of electron-positron annihilation, their energy density after their decoupling would be equal to $3\times (7/8)\times (4/11)^{4/3}\,\rho_{\rm cmb}$, with the first factor capturing the contributions from the 3 active species; the second the difference between fermions and bosons; and the last the relative heating of the photons in the CMB by electron-positron annihilation. However decoupling is not a discrete event and occurs close to the time of electron-positron annihilation, so the neutrinos share a bit in the heating, with the factor of 3 replaced by $N_{\rm eff}=3.046$. The additional fraction depends not only on well-known neutrino scattering rates but also finite temperature quantum corrections. 
Upcoming experiment have the potential to measure this tiny deviation of $N_{\rm eff}$ from 3. This will be an amazing test of our understanding of the Universe when it was about a second old.
Furthermore any significant deviation from this value could be a hint of a different scenario not predicted by the standard model. 
%==========
\begin{figure}[htbp]
\begin{center}
\includegraphics{prior_massless_neutrinos_snow_mass.pdf}
\caption{Projected 1-sigma error on $N_{\rm eff}$ as a function of the strength of the prior on the other 6 cosmological parameters of the standard $\Lambda$CDM model. Each prior is expressed in units of the error that would be obtained internally from our suite of CMB experiments; e.g., $\sigma(H_0)_{\rm old}=4.1\%$. The parameters that would be most useful to constrain externally for the purposes of determining $N_{\rm eff}$ are $n_s$ and $\Omega_bh^2$. The bottom-most (grey) curve shows the impact of imposing the given prior on {\it all} the other parameters simultaneously.}
\label{fig:prior_massless_neutrinos}
\end{center}
\end{figure}
%==========
\reffig{prior_massless_neutrinos} shows projections for how well our suite of CMB experiments will do at measuring $N_{\rm eff}$ as a function of priors on the other 6 parameters in the standard cosmological model ($H_0, \Omega_mh^2,\Omega_bh^2, n_s, A_s,$ and $\tau$). With no priors on the other parameters, the projected $1$-sigma error is  about $0.02$, suggesting a 2-sigma detection of the deviation from $N_{\rm eff}=3$. The sensitivity comes from the effect of extra species on the damping tail of the CMB anisotropies, both in temperature and polarization, so parameters that also affect the shape of the damping tail, like the slope $n_s$ and the baryon density $\Omega_bh^2$, are most degenerate with $N_{\rm eff}$. As such, \reffig{prior_massless_neutrinos} shows that obtaining external priors on either of these would reduce the errors on $N_{\rm eff}$ to ensure a 3-sigma detection of the partial decoupling prediction. The bottom-most curve shows that if all parameters were constrained externally, then the projected error on $N_{\rm eff}$ would improve significantly.

\subsection{Neutrino Masses}

@@

\subsection{Dark Energy Equation of State}

@@

\reffig{prior_w} shows that without any priors, CMB experiments will not do much better than current constraints, which hover around 10\%. However, an external constraint on the Hubble constant would improve the CMB constraints on $w$ considerably. {\bf We should try to understand why.}

%{\bf I don't really understand this table.}
%\begin{table}[htdp]
%
%\begin{center}
%\begin{tabular}{|c|c|c|c|c|c|c|}
%\hline
%$H_{0}$ &$ M_{\nu}$ &$\Omega_{bc}h^{2}$&$\Omega_{b}h^{2}$&$\tau$&$A_{s}$&$n_{s}$ \\
%$0.80 \%$&$40.92\%$&$0.44\%$&$0.11\%$&$3.075\%$&$0.53\%$&$0.18\%$\\
%
%\hline
%\end{tabular}
%\caption{How well we do constrain separate parameters with this data without any external prior? Things to notice: this is done with the Zhen test parameters ($\ell<3000$ and no error on lensing). Now if we trust it CMB alone can get a $0.8\%$ error on $H_{0}$ thus I am not surprised if a prior on H would not help. However what do we think about it? is it really CMB better than SN. People will not agree on that. $\tau$ will probably improve and also M$\nu$ from BAO may help improving parameters.}
%\end{center}
%\label{default}
%\end{table}%

>>>>

% Group addresses by affiliation; use superscriptaddress for long
% author lists, or if there are many overlapping affiliations.
% For Phys. Rev. appearance, change preprint to twocolumn.
% Choose pra, prb, prc, prd, pre, prl, prstab, prstper, or rmp for journal
%  Add 'draft' option to mark overfull boxes with black boxes
%  Add 'showpacs' option to make PACS codes appear
%  Add 'showkeys' option to make keywords appear
\documentclass[aps,prd,reprint,superscriptaddress]{revtex4-1}
%\documentclass[aps,prl,preprint,superscriptaddress]{revtex4-1}
%\documentclass[aps,prl,reprint,groupedaddress]{revtex4-1}
%\usepackage{bm}
\usepackage{amsmath}
\usepackage{lineno}
\usepackage{graphicx}

%\newcommand\refsec[1]{\S\ref{sec:#1}}
%\newcommand\refeq[1]{Eq.~(\ref{eqn:#1})}
%\input{papers_packages_command._include_tex}
\newcommand\refsec[1]{\S\ref{sec:#1}}
\newcommand\refeq[1]{Eq.~(\ref{eqn:#1})}
\newcommand{\refssec}[1]{section~\ref{subsec:#1}}
\newcommand{\reffig}[1]{Fig.~\ref{fig:#1}}
%\newcommand{\apj}{ApJ}
\newcommand{\physrep}{Physics Reports}
\newcommand{\aapr}{A\&A Rev.}
\newcommand{\apjl}{ApJL}
\newcommand{\apjs}{ApJS}
\newcommand{\aap}{A\&A}
\newcommand{\mnras}{MNRAS}
%\newcommand{\prd}{Phys. Rev. D}
\newcommand{\physrev}{Phys. Rev.}
\newcommand{\physrevlett}{Phys. Rev. Lett.}
\newcommand{\jcap}{{\em JCAP }}
\newcommand{\aj}{AJ }
\bibliographystyle{apsrev4-1}

@@

\date{\today}
\begin{abstract}
The next generation of cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, Dark Matter and Dark Energy. 
Indeed the low level of noise, together with improved angular resolution, will drastically increase the signal to noise of the CMB polarized signal as well as the reconstructed lensing potential of high redshift large scale structure. 
Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments. Here, we examine quantitatively, with a Fisher matrix approach, the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 experiment.
We find that CMB S4... 
\end{abstract}

@@

@@

To estimate errors on cosmological parameters we follow the Bayes theorem that relates the likelihood to measure a set of data given the parameters of the model $\mathcal{L}(d|\theta)$, to what we want: the posterior probability of those parameters given the data, $\mathcal{L}(\theta|d)$.
These two are related by the prior probability of the parameters $P(\theta)$ through:
\begin{equation}
\mathcal{L}(\theta|d)\propto \mathcal{L}(d|\theta)P(\theta),
\end{equation}
where, as usual, we have neglected the probability of the data $P(d)$.
These general framework can be simplified in our case. We will focus not on an entirely general posterior distribution, but we will assume that the likelihoods are gaussian.
Furthermore we will not derive the global shape of the likelihood but we will obtain parameters errors by studying small perturbations around its maximum. These are the two basic assumptions of the Fisher-matrix approach.

@@

In this subsection we describe the assumptions we made in calculating the elements that goes in \refeq{Fij_def2} and, in particular, the power spectra $C_{\ell}$ and the noise power $N_{\ell}$.
Firstly we parametrize our cosmology using a flat $\nu \Lambda$CDM universe. We allow a set different Dark Energy models by introducing the equations of state parameter $w$ as a varying parameter.
We chose our fiducial parameters following Table 2 of \textit{Planck} best fit \cite{planck-collaboration:2014g}, i.e. $\Omega_c h^2 = 0.12029$, $\Omega_b h^2 = 0.022068$, $A_s = 2.215\times10^{-9}$ at $k_0 = 0.05\ {\rm Mpc}^{-1}$, $n_s = 0.9624$, $\tau = 0.0925$, and $H_0 = 67.11$ km/s/Mpc. Regarding the $\Lambda$CDM extension, we chose a neutrino energy densities $\Omega_{\nu} h^2$=0.0009, which corresponds to $M_{\nu}$ $\simeq$ 85\ meV, a standard $N_{\rm eff}=3.046$ and a dark energy equation of state, $w=-1$.
Given the model, we use CAMB \cite{Lewis:1999bs} to compute the power spectra $C_{\ell}$ at the fiducial values and at those needed to compute derivatives using \refeq{deriv}. Notice that while we vary one parameter in \refeq{deriv} we keep all the others fixed with the exception of $\Omega_{\Lambda}$ which is always changed in order to keep the universe flat ($\Omega_{\rm k} = 0$).

@@

\section{Results \label{sec:results}}
In this section we present our main results. We will start from..

@@

In the standard cosmology, three active neutrinos are thermally produced in the early universe. Were they to decouple well before the epoch of electron-positron annihilation, their energy density after their decoupling would be equal to $3\times (7/8)\times (4/11)^{4/3}\,\rho_{\rm cmb}$, with the first factor capturing the contributions from the 3 active species; the second the difference between fermions and bosons; and the last the relative heating of the photons in the CMB by electron-positron annihilation. However decoupling is not a discrete event and occurs close to the time of electron-positron annihilation, so the neutrinos share a bit in the heating, with the factor of 3 replaced by $N_{\rm eff}=3.046$. The additional fraction depends not only on well-known neutrino scattering rates but also finite temperature quantum corrections. 
Upcoming experiment have the potential to measure this tiny deviation of $N_{\rm eff}$ from 3. This will be an amazing test of our understanding of the Universe when it was about a second old.
Furthermore any possible significant deviation from this value could be a hint of a different scenario not predicted by the standard model. 
%==========
\begin{figure}[htbp]
\begin{center}
\includegraphics{prior_massless_neutrinos_snow_mass.pdf}
\caption{Projected 1-sigma error on $N_{\rm eff}$ as a function of the strength of the prior on the other 6 cosmological parameters of the standard $\Lambda$CDM model. Each prior is expressed in units of the error that would be obtained internally from our suite of CMB experiments; e.g., $\sigma(H_0)_{\rm old}=4.1\%$. The parameters that would be most useful to constrain externally for the purposes of determining $N_{\rm eff}$ are $n_s$ and $\Omega_bh^2$. The bottom-most (grey) curve shows the impact of imposing the given prior on {\it all} the other parameters simultaneously.}
\label{fig:prior_massless_neutrinos}
\end{center}
\end{figure}
%==========

\reffig{prior_massless_neutrinos} shows projections for how well our suite of CMB experiments will do at measuring $N_{\rm eff}$ as a function of priors on the other 6 parameters in the standard cosmological model ($H_0, \Omega_mh^2,\Omega_bh^2, n_s, A_s,$ and $\tau$). With no priors on the other parameters, the projected $1$-sigma error is  about $0.02$, suggesting a 2-sigma detection of the deviation from $N_{\rm eff}=3$. The sensitivity comes from the effect of extra species on the damping tail of the CMB anisotropies, both in temperature and polarization, so parameters that also affect the shape of the damping tail, like the slope $n_s$ and the baryon density $\Omega_bh^2$, are most degenerate with $N_{\rm eff}$. As such, \reffig{prior_massless_neutrinos} shows that obtaining external priors on either of these would reduce the errors on $N_{\rm eff}$ to ensure a 3-sigma detection of the partial decoupling prediction. The bottom-most curve shows that if all parameters were constrained externally, then the projected error on $N_{\rm eff}$ would improve significantly.

\subsection{Neutrino Masses, $\sum m_\nu$}

@@

\subsection{Dark Energy Equation of State, $w$}

@@

\reffig{prior_w} shows that without any priors, CMB experiments will not do much better than current constraints, which hover around 10\%. However, an external constraint on the Hubble constant would improve the CMB constraints on $w$ considerably. {\bf We should try to understand why. Probably just a geometrical degeneracy in the way $w$ and $H_{0}$ enter the definition of the distance to the LSS $D(z_{\rm LSS})$}

\section{Conclusions  \label{sec:conclusions}}
